---
title: "Lab 12 - Smoking during pregnancy"
author: "Allison Li"
date: "20250430"
output: github_document
---

### Load packages and data

```{r load-packages, message=FALSE}
##install.packages("openintro")
##install.packages("infer")
library(infer)
library(tidyverse) 
library(tidymodels)
library(openintro)

##set a seed
set.seed(1234)
```

### Exercise 1

```{r outliers}
data(ncbirths)

boxplot(ncbirths$mage, main = "of Mother's Age", ylab = "Age")
boxplot(ncbirths$fage, main = "Father's Age", ylab = "Age")
boxplot(ncbirths$weeks, main = "length of pregnancy", ylab = "Weeks")
boxplot(ncbirths$visits, main = "number of hospital visits during pregnancy", ylab = "Times")
boxplot(ncbirths$gained, main = "weight gained by mother during pregnancy", ylab = "Pounds")
boxplot(ncbirths$weight, main = "weight of the baby at birth", ylab = "Pounds")

```

Categorical variables:mature, premie, marital, lowbirthweight, gender, habit, whitemom
Numeric variables:fage, mage, weeks, visits, gained, weight
There is one outlier for mage, two for fage, multiple (11) for weeks, 7 for visits, multiple for gained, and many outliers for weight.

The cases are the comprehensive birth record for infants in North Carolina, 2004. Each case in the dataset is the information about their father and mother during pregancy for each infant. There are 1000 cases in the sample

### Exercise 2

```{r}
ncbirths_white <- ncbirths %>%
  filter(whitemom == "white")

mean(ncbirths_white$weight, na.rm = TRUE)
```

### Exercise 3

I believe the criteria are satisfied. The dataset has relatively large sample (1000) and the observations are random and independent of each other.

### Exercise 4

```{r bootstrap}
set.seed(1234)
bootstrap <- ncbirths_white %>%
  specify(response = weight) %>%
  generate(reps = 1500, type = "bootstrap") %>%
  calculate(stat = "mean")
summary(bootstrap)

obs_mean <- mean(ncbirths_white$weight)
centered_dist <- bootstrap %>%
  mutate(stat = stat - obs_mean + 7.43)

##I asked gpt bc I do not know how to calculate the p value in this way
p_value <- centered_dist %>%
  summarise(
    p_val = mean(abs(stat - 7.43) >= abs(obs_mean - 7.43))
    ) %>%
  pull(p_val)

bootstrap %>% 
  summarize(lower = quantile(stat, .025),
            uppter = quantile(stat, .975))

##Plot
ggplot(data = centered_dist, mapping = aes(x = stat)) +
  geom_histogram(binwidth = .01, fill = "skyblue", color = "pink") +
  geom_vline(xintercept = obs_mean, color = "red", linetype = "dashed", size = 1) +
  labs(title = "Null distribution of means")

```

The null hypothesis: H0: u = 7.43; 
Alternative hypothesis: H1: u ≠ 7.43

According to the table, there is no sample with a mean less than 7.25.
According to the output, the observed mean birth weight of babies born to White mothers in the sample was 7.25 pounds. To test the hypothesis, we generated 150 bootstrap samples of the data. The resulting p-value was .002, which suggests we can reject the null hypothesis and conclude that the observed value is significantly lower than the historical weight of 7.43 pounds. 

### Exercise 5

```{r habit and weight}
ggplot(data = ncbirths, aes(x = habit, y = weight)) +
  geom_boxplot(fill = "pink") +
  labs(title = "Baby Weight by Mother's Smoking Habit",
       x = "Mother's Smoking Habit",
       y = "Baby Weight (lbs)")
```

Based on the graph, we can assume that mother who smoke tends to have a lighter baby weight, although we do not know if the difference is statistically significant (since there are a lot of observed outliers in the nonsmoker group). 

### Exercise 6

```{r cleaned version}
ncbirths_clean <- ncbirths %>%
  filter(!is.na(habit) & !is.na(weight))
```

### Exercise 7

```{r}
ncbirths_clean %>%
  group_by(habit) %>%
  summarize(mean_weight = mean(weight))
```

### Exercise 8

H0: the weight of the infant whose mother is smoker = the weight of the infant whose mother does not smoke (μ1 = μ2)

H1: the weight of the infant whose mother is smoker ≠ the weight of the infant whose mother does not smoke (μ1 ≠ μ2)

### Exercise 9

```{r ttest}
t.test(weight ~ habit, data = ncbirths_clean)
```

according to the output, p = .02, which is statistically significant at the .05 significance level. Therefore, it can be conclude that the average weight of infants whose mother is not smoker is significantly heavier than the average weight of infants whose mother is smoker. Additionally, we are 95% confident that the true difference in means between non-smoker and smoker mother is between about 0.05 and 0.58 pounds. 

### Exercise 10

```{r CI}
set.seed(1234)
bootstrap_clean <- ncbirths_clean %>%
  specify(weight ~ habit) %>%
  generate(reps = 1500, type = "bootstrap") %>%
  calculate(stat = "diff in means", order = c("nonsmoker", "smoker"))
bootstrap_clean %>%
  get_confidence_interval(level = 0.95, type = "percentile")
```

The bootstrapping showed a slightly different value for CI than the sample, but it still indicates a statistically significant difference between the two groups.

### Exercise 11

```{r mage cutoff}
ncbirths %>%
  summarise(
    min = min(mage, na.rm = TRUE),
    q1 = quantile(mage, 0.25, na.rm = TRUE),
    median = median(mage, na.rm = TRUE),
    mean = mean(mage, na.rm = TRUE),
    q3 = quantile(mage, 0.75, na.rm = TRUE),
    max = max(mage, na.rm = TRUE),
    sd = sd(mage, na.rm = TRUE)
  )

```

I looked at the max, min, and mean of mother's age, and decided to set the cutoff for younger and mature mother at the age of 27, which is the mean and median (lucky they are the same number). 

### Exercise 12

H0: the proportion of low birth weight babies is the same for mature mothers and younger mother (μ1 = μ2);

H1: the proportion of low birth weight babies is significantly higher for mature mothers than younger mother (μ1 > μ2)

```{r low birth weight babies and mother age}
set.seed(1234)
ncbirths_lowbirth <- ncbirths %>%
  mutate(lowbirthweight = if_else(lowbirthweight == "low", 1, 0)) %>%
  mutate(age_group = if_else(mage < 27, "younger", "older"))

## This would tell me the observed difference in low birthweight proportions
obs_diff <- ncbirths_lowbirth %>%
  group_by(age_group) %>%
  summarise(prop_low = mean(lowbirthweight), .groups = "drop") %>%
  pivot_wider(names_from = age_group, values_from = prop_low) %>%
  mutate(diff = older - younger) %>%
  pull(diff)

## The test
ncbirths_lowbirth <- ncbirths_lowbirth %>%
  mutate(lowbirthweight_str = as.character(lowbirthweight))

lowbirthboot <- ncbirths_lowbirth %>%
  specify(lowbirthweight_str ~ age_group, success = "1") %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1500, type = "permute") %>%
  calculate(stat = "diff in props", order = c("younger", "older"))

# One-sided p-value from GPT 
p_value <- lowbirthboot %>%
  get_p_value(obs_stat = obs_diff, direction = "right")

p_value

```

Based on the results, a p-value of .41 is found, which is larger than the significance level of 0.05. Therefore, we failed to reject the null hypothesis, which indicates that there is no statistically significance between  the proportion of low birth weight babies from older mothers (27 and above) and among younger mothers.

### Exercise 13

```{r CI mothers}
ci <- lowbirthboot %>%
  get_confidence_interval(level = 0.95, type = "percentile")

ci

```
Based on 1500 bootstrap samples, the 95% CI is (-.04, 0.04).Since the interval includes 0, we can conclude that there no statistically significant difference in low birth weight rates between the two age groups of mothers.
